---
title: "Report"
output: html_document
---


First, lets get the data clean.
```{r echo=FALSE, cache=TRUE}
# DOwnloading the data sets if necessary
if(!file.exists("pml-training.csv"))
{
        download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-training.csv", "curl")
}
if(!file.exists("pml-testing.csv"))
{
        download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-testing.csv", "curl")
}
# Loading the data sets
training <- read.csv("pml-training.csv")
testing <- read.csv("pml-testing.csv")
# Deleting sumarization variables in the training data
training <- training[,-grep("kurtosis", names(training))]
training <- training[,-grep("skewness", names(training))]
training <- training[,-grep("max", names(training))]
training <- training[,-grep("min", names(training))]
training <- training[,-grep("amplitude", names(training))]
training <- training[,-grep("var", names(training))]
training <- training[,-grep("avg", names(training))]
training <- training[,-grep("stddev", names(training))]
# Deleting sumarization variables in the testing data
testing <- testing[,-grep("kurtosis", names(testing))]
testing <- testing[,-grep("skewness", names(testing))]
testing <- testing[,-grep("max", names(testing))]
testing <- testing[,-grep("min", names(testing))]
testing <- testing[,-grep("amplitude", names(testing))]
testing <- testing[,-grep("var", names(testing))]
testing <- testing[,-grep("avg", names(testing))]
testing <- testing[,-grep("stddev", names(testing))]
# Deleting timestamp data from both data sets
training <- training[,-grep("timestamp", names(training))]
testing <- testing[,-grep("timestamp", names(testing))]
# Deleting window data from both data sets
training <- training[,-grep("window", names(training))]
testing <- testing[,-grep("window", names(testing))]
# Deleting X variable from both data sets
training <- training[,-1]
testing <- testing[,-1]
# Deleting User Name variable from both data sets
training <- training[,-1]
testing <- testing[,-1]
# Factor total_accel variables
training$total_accel_belt <- factor(training$total_accel_belt)
training$total_accel_arm <- factor(training$total_accel_arm)
training$total_accel_dumbbell <- factor(training$total_accel_dumbbell)
training$total_accel_forearm <- factor(training$total_accel_forearm)

testing$total_accel_belt <- factor(testing$total_accel_belt)
testing$total_accel_arm <- factor(testing$total_accel_arm)
testing$total_accel_dumbbell <- factor(testing$total_accel_dumbbell)
testing$total_accel_forearm <- factor(testing$total_accel_forearm)
```


Now lets gets some cross validation data sets.
```{r, echo=FALSE, cache=TRUE}
library(caret)
set.seed(32343)

# Let's create our cross validation data sets
inTrain <- createDataPartition(y=training$classe,
                              p=0.60, list=FALSE)
train <- training[inTrain,]
testTemp <- training[-inTrain,]
inTest <- createDataPartition(y=testTemp$classe,
                              p=0.50, list=FALSE)
test <- testTemp[-inTest,]
validate <- testTemp[-inTest,]
```

At this point lets reduce the dimensionality of our data set using PCA.

```{r, echo=FALSE, cache=TRUE}
tempNames <- names(train)[-grep("total_accel", names(train))]
preProcBelt <- preProcess(train[,tempNames[grep("_belt", tempNames)]],method="pca", thresh = 0.80)
preProcForearm <- preProcess(train[,tempNames[grep("_forearm", tempNames)]],method="pca", thresh = 0.80)
preProcDumbbell <- preProcess(train[,tempNames[grep("_dumbbell", tempNames)]],method="pca", thresh = 0.80)
preProcArm <- preProcess(train[,c(tempNames[grep("_arm_", tempNames)], tempNames[grep("_arm$", tempNames)])],method="pca", thresh = 0.80)

```

- The variables measured by the Belt Sensor are `r preProcBelt$dim[2]`. And the number of Principal Components are `r preProcBelt$numComp`.
- The variables measured by the Belt Sensor are `r preProcForearm$dim[2]`. And the number of Principal Components are `r preProcForearm$numComp`.
- The variables measured by the Belt Sensor are `r preProcDumbbell$dim[2]`. And the number of Principal Components are `r preProcDumbbell$numComp`.
- The variables measured by the Belt Sensor are `r preProcArm$dim[2]`. And the number of Principal Components are `r preProcArm$numComp`.

Now, let's use our new components to generate new training data using that dimensions:

```{r, echo=FALSE, cache=TRUE}
trainBelt <- predict(preProcBelt,train[,tempNames[grep("_belt", tempNames)]])
trainForearm <- predict(preProcForearm,train[,tempNames[grep("_forearm", tempNames)]])
trainDumbbell <- predict(preProcDumbbell,train[,tempNames[grep("_dumbbell", tempNames)]])
trainArm <- predict(preProcArm,train[,c(tempNames[grep("_arm_", tempNames)], tempNames[grep("_arm$", tempNames)])])
```

Let's fit a model using "rf":

```{r, echo=FALSE, cache=TRUE}
trainAllPCA <- data.frame(trainArm, trainDumbbell, trainForearm, trainBelt, classe=train$classe)
modFitAllPCA <- train(classe ~ ., data=trainAllPCA, method="rf")
```

Now let's test our model with the cross validation data:

```{r, echo=FALSE, cache=TRUE}
testBelt <- predict(preProcBelt, test[,tempNames[grep("_belt", tempNames)]])
testForearm <- predict(preProcForearm, test[,tempNames[grep("_forearm", tempNames)]])
testDumbbell <- predict(preProcDumbbell, test[,tempNames[grep("_dumbbell", tempNames)]])
testArm <- predict(preProcArm, test[,c(tempNames[grep("_arm_", tempNames)], tempNames[grep("_arm$", tempNames)])])

testAllPCA <- data.frame(testArm, testDumbbell, testForearm, testBelt, classe=test$classe)
testPredAllPCA <- predict(modFitAllPCA,testAllPCA)
confMtxTestAllPCA <- confusionMatrix(test$classe, testPredAllPCA)

validateBelt <- predict(preProcBelt, validate[,tempNames[grep("_belt", tempNames)]])
validateForearm <- predict(preProcForearm, validate[,tempNames[grep("_forearm", tempNames)]])
validateDumbbell <- predict(preProcDumbbell, validate[,tempNames[grep("_dumbbell", tempNames)]])
validateArm <- predict(preProcArm, validate[,c(tempNames[grep("_arm_", tempNames)], tempNames[grep("_arm$", tempNames)])])

validateAllPCA <- data.frame(validateArm, validateDumbbell, validateForearm, validateBelt, classe=validate$classe)
validatePredAllPCA <- predict(modFitAllPCA,validateAllPCA)
confMtxValidateAllPCA <- confusionMatrix(validate$classe, validatePredAllPCA)
```


```{r, echo=FALSE, cache=TRUE}
```


